A Mechanised Library for Generic (Co)Programming

### 2024-03-18 paper/paper.tex
RW

\subsection{Recursive Anamorphisms}
Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
inductive data type in the following way.  Intuitively, a recursive coalgebra
can be applied only finitely many times, therefore when this is applied to an
anamorphism the only possible traces it can produce from the seed function are
the finite ones. We denote this special kind of recursion scheme \emph{recursive
anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
from the universality property of the hylomorphism and the fact that recursive
anamorphisms satisfy the same equation.

We now show how this is proven in the Coq development. First we define a
recursive anamorphism as a function into \coq{LFix F}:
\begin{coqcode}
Definition rana_f__ `{setoid A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).

Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
\end{coqcode}
We omit the boilerplate details to obtain the setoid morphism.
The recursive anamorphism has the following universality property:
\begin{coqcode}
Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
    : f =e rana h <-> f =e l_in \o fmap f \o h.
\end{coqcode}
We can now show that every hylomorphism is composed by a recursive anamorphism
followed by a catamorphism:
\begin{coqcode}
Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
  (g : Alg F B) (h : RCoalg F A)
: cata g \o rana h =e hylo g h.
\end{coqcode}
This result will allow us to \emph{calculate} programs by providing small, easy
to verify implementations, and combining them into efficient recursive functions
by using Coq's proofs of hylomorphism laws as optimisation passes.

@@

Abreu et al.\ propose\ldots~\cite{AbreuDHJMS23}. Their approach has similarities to what is presented here.

They differ in requiring the assumption of the functional extensionality axiom.
Additionally, they are concerned with the erasure of termination proofs but have
little discussion of what the resulting extracted code \emph{looks like} (that
is, whether the extracted OCaml code resembles the natural formulation of the
recursive function).
\mvol{We should find out for sure if they use \texttt{obj.magic}.}

>>>>

% \subsection{Recursive Anamorphisms}
% Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
% inductive data type in the following way.  Intuitively, a recursive coalgebra
% can be applied only finitely many times, therefore when this is applied to an
% anamorphism the only possible traces it can produce from the seed function are
% the finite ones. We denote this special kind of recursion scheme \emph{recursive
% anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
% can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
% exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
% algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
% from the universality property of the hylomorphism and the fact that recursive
% anamorphisms satisfy the same equation.
% 
% We now show how this is proven in the Coq development. First we define a
% recursive anamorphism as a function into \coq{LFix F}:
% \begin{coqcode}
% Definition rana_f__ `{setoid A} (c : Coalg F A)
%   : forall x : A, RecF c x -> LFix F
%   := fix f (x : A) (H : RecF c x) :=
%        let hx := c x in
%        LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
% 
% Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
% \end{coqcode}
% We omit the boilerplate details to obtain the setoid morphism.
% The recursive anamorphism has the following universality property:
% \begin{coqcode}
% Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
%     : f =e rana h <-> f =e l_in \o fmap f \o h.
% \end{coqcode}
% We can now show that every hylomorphism is composed by a recursive anamorphism
% followed by a catamorphism:
% \begin{coqcode}
% Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
%   (g : Alg F B) (h : RCoalg F A)
% : cata g \o rana h =e hylo g h.
% \end{coqcode}
% This result will allow us to \emph{calculate} programs by providing small, easy
% to verify implementations, and combining them into efficient recursive functions
% by using Coq's proofs of hylomorphism laws as optimisation passes.

@@

Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing.  This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}.
The authors justify well the use of impredicative \coq{Set} and its
compatibility with the functional extensionality axiom. In contrast, our
development remains entirely \emph{axiom-free}. Another key difference with our
approach is that they do not discuss what the resulting extracted code
\emph{looks like} (that is, whether the extracted OCaml code resembles the
natural formulation of the recursive function). Through experiments, we found
that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
structure that may be hard to understand or interface with other OCaml code.
Due to the great benefit of entirely avoiding termination proofs, it would be
interesting to extend their approach to improve code extraction.
